Use and acquisition of causal relations for decision making under uncertainty using imperfect information games

### 2020-09-02 3rd_year_eval/Slides/3rd_year_slides.tex
quitando research problem

\subsection{Research Problem}
\begin{frame}
\frametitle{Research Problem}
We recall here the Research Problem described in \cite{gonzalez2019causal}:
\begin{itemize}
\item Let  $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem under uncertainty whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. We ask how a decision maker could \textit{learn} about the causal structure that controls his environment in order to \textit{make good choices} with respect to his preferences.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Second Year}
\begin{itemize}
\item Using the previous inequality we were able to define the notion of a Causal Game and the idea of a Nash equilibria within such game. We define a function $u^C_i : A \to \mathbb{R}^{+}$, for each player:
\begin{equation}{\label{games_causal_ut}}
u^C_i (a) = \sum_{f \in A}  u_i(a) \left( \sum_g P^g_i (h_i(f) | do(a_i), a_{-i}) P_C(g) \right).
\end{equation}
\item We define a Nash equilibrium for this \textit{causal strategic game} is an action profile $a^\ast \in A$ if and only if
\begin{equation}{\label{nash_eq}}
 u^C_i(a^\ast) \geq u^C_i(a_i, a^\ast_{-i}) \textrm{ for any other } a_i \in A_i. 
 \end{equation}
 \item We argued that Reinforcement Learning is not a Causal Problem in itself in \cite{gonzalez2019reinforcement}.
 \item Technical report CCC-19-002 which we cite here as \cite{gonzalez2019causal}. 
\end{itemize}
\end{frame}

@@

\subsection{Causal Games}
\begin{frame}
\frametitle{Causal Games: New work}
We improved the work done in the Second Year by modifying Equation \ref{games_causal_ut} in the following way:
\begin{itemize}
\item For each player $i \in N$ in the strategic game, we define the following probability distribution over consequences:
\begin{equation}{\label{causal_utility}}
p^a_i (c) =  p^\omega_i (c | do(a_i), a_{-i}) p_i(\omega)\textrm{ for } a \in A=A_1 \times \cdots \times A_N.
\end{equation}
\item We now define:
\begin{equation}
u^C_i (a) = \sum_{c \in C}  u_i(c) p^a_i (c) \textrm{ for } a \in A=A_1 \times \cdots \times A_N.
\end{equation}
\item A Nash equilibrium for this \textit{causal strategic game} is an action profile $a^\ast \in A$ if and only if
\begin{equation}
 u^C_i(a^\ast) \geq u^C_i(a_i, a^\ast_{-i}) \textrm{ for any other } a_i \in A_i. 
 \end{equation}
\item This work, \cite{gonzalez2019games}, was presented at a MICAI 2019 Workshop.
\end{itemize}
\end{frame}

\subsection{Causal-Structure Learning: A Random Graph Approach}{\label{random_graph}}
\begin{frame}
\frametitle{Causal-Structure Learning: A Random Graph Approach}
\begin{itemize}
\item Human beings focus on \textit{local} aspects while learning causal relations which are later unified into a single structure \citep{fernbach2009causal,waldmann2008causal, danks2014unifying}.
\item Following this idea, \cite{wellen2012learning} propose a model to explain how observations  and interventions are used by human beings to learn causal relations in terms of a local prediction-error learning.
\item Following this line of thought, we propose here a local probabilistic encoding of the uncertainty that a decision maker has over the existence or not of causal relations between variables.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Random Graphs}
\begin{itemize}
\item What is a Random Graph? 
\item Were invented by Erdös and Renyi. 
\item They are random objects which take its values in the space of graphs (or adjacency matrices).
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Random Graphs: most common models}
\begin{itemize}
\item We will be using the following model:
\item Among $N$ vertices, draw an edge between vertices $i$ and $j$ with probability $p_{ij}$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Assumptions}
\begin{itemize}
\item Rational agent.
\item Causal sufficiency
\item Knowledge of the \textit{variables} of the model.
\item Convergence to true parameter of Bayesian estimators \cite{autzen2018bayesian}.
\item Existence of a causal ordering of the variables.
\end{itemize}
\end{frame}

@@

\subsection{Implementation}
\begin{frame}
\frametitle{Implementation}
\begin{itemize}
\item We are using the same setting as in the experiments described in Year 1; this is, a patient sick with an unknown disease and a doctor who must choose between two possible treatments, each one with some risk of causing a life-threatening allergic reaction. Our agent (the doctor) must now learn which variables (disease, treatment, reaction, dying) are causally related.
\end{itemize}
\begin{enumerate}
\item Initialize $p_{ij}$ randomly
\item Initialize graph $G$ with link from node $i$ to node $j$ with probability $p_{ij}$
\item Find optimal action $a^\ast$ for graph $G$ and probabilities given by a count of the observations.
\item Calculate probability of the approximate model $G$ producing the full observation $X_1 = a^\ast, X_2=x_2,...,X_n=x_n$. 
\item Update $p_{ij}^{t+1} \propto p(X_1 = a^\ast, X_i = x_i, X_j=x_j, X_n=x_n | \textrm{subgraph formed by nodes 1,i,j,n})p_{ij}^t.$
\end{enumerate}
\end{frame}

>>>>

%\subsection{Research Problem}
%\begin{frame}
%\frametitle{Research Problem}
%We recall here the Research Problem described in \cite{gonzalez2019causal}:
%\begin{itemize}
%\item Let  $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem under uncertainty whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. We ask how a decision maker could \textit{learn} about the causal structure that controls his environment in order to \textit{make good choices} with respect to his preferences.
%\end{itemize}
%\end{frame}

@@

\begin{frame}
\frametitle{Second Year}
\begin{itemize}
\item Using the previous inequality we were able to define the notion of a Causal Game and propose a first version of a Nash equilibria within such game. We define a function $u^C_i : A \to \mathbb{R}^{+}$, for each player:
\begin{equation}{\label{games_causal_ut}}
u^C_i (a) = \sum_{f \in A}  u_i(a) \left( \sum_g P^g_i (h_i(f) | do(a_i), a_{-i}) P_C(g) \right).
\end{equation}
\item We define a Nash equilibrium for this \textit{causal strategic game} is an action profile $a^\ast \in A$ if and only if
\begin{equation}{\label{nash_eq}}
 u^C_i(a^\ast) \geq u^C_i(a_i, a^\ast_{-i}) \textrm{ for any other } a_i \in A_i. 
 \end{equation}
 \item We argued that Reinforcement Learning is not a Causal Problem in itself in \cite{gonzalez2019reinforcement}.
 \item Technical report CCC-19-002 which we cite here as \cite{gonzalez2019causal}. 
\end{itemize}
\end{frame}

@@

\subsection{Causal Games}
\begin{frame}
\frametitle{Causal Games: New work}
We continued the work done in the Second Year by modifying Equation \ref{games_causal_ut} in the following way:
\begin{itemize}
\item For each player $i \in N$ in the strategic game, we define the following probability distribution over consequences:
\begin{equation}{\label{causal_utility}}
p^a_i (c) =  p^\omega_i (c | do(a_i), a_{-i}) p_i(\omega)\textrm{ for } a \in A=A_1 \times \cdots \times A_N.
\end{equation}
\item We now define:
\begin{equation}
u^C_i (a) = \sum_{c \in C}  u_i(c) p^a_i (c) \textrm{ for } a \in A=A_1 \times \cdots \times A_N.
\end{equation}
\item A Nash equilibrium for this \textit{causal strategic game} is an action profile $a^\ast \in A$ if and only if
\begin{equation}
 u^C_i(a^\ast) \geq u^C_i(a_i, a^\ast_{-i}) \textrm{ for any other } a_i \in A_i. 
 \end{equation}
\item This work, \cite{gonzalez2019games}, was presented at a MICAI 2019 Workshop.
\end{itemize}
\end{frame}

\subsection{Causal-Structure Learning: A Random Graph Approach}{\label{random_graph}}
\begin{frame}
\frametitle{Causal-Structure Learning: A Random Graph Approach}
\begin{itemize}
\item Human beings focus on \textit{local} aspects while learning causal relations which are later unified into a single structure \citep{fernbach2009causal,waldmann2008causal, danks2014unifying}.
\item Following this idea, \cite{wellen2012learning} propose a model to explain how observations  and interventions are used by human beings to learn causal relations in terms of a local prediction-error learning.
\item We propose here a local probabilistic encoding of the uncertainty that a decision maker has over \textit{the existence or not of causal relations} between variables.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Random Graphs}
\begin{itemize}
\item What is a Random Graph? 
\item Were invented by Erdös and Renyi. 
\item They are random objects which take its values in the space of graphs (or adjacency matrices).
\item For example, fix a set of nodes, and connect two nodes with some probability $p$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Assumptions}
\begin{itemize}
\item Rational agent.
\item Causal sufficiency.
\item Knowledge of the \textit{variables} of the model.
\item Convergence to true parameter of Bayesian estimators \cite{autzen2018bayesian}.
\item Existence of a causal ordering of the variables.
\end{itemize}
\end{frame}

@@

\subsection{Implementation}
\begin{frame}
\frametitle{Implementation}
\begin{itemize}
\item We are using the same setting as in the experiments described in Year 1; this is, a patient sick with an unknown disease and a doctor who must choose between two possible treatments, each one with some risk of causing a life-threatening allergic reaction. Our agent (the doctor) must now learn which variables (disease, treatment, reaction, dying) are causally related.
\end{itemize}
\begin{enumerate}
\item Initialize $p_{ij}$ randomly
\item Initialize graph $G$ with link from node $i$ to node $j$ with probability $p_{ij}$
\item Find optimal action $a^\ast$ for graph $G$ and probabilities given by a count of the observations.
\item Observe the full observation $X_1 = a^\ast, X_2=x_2,...,X_n=x_n$. 
\item Using current $G$, update $p_{ij}^{t+1} \propto p(X_1 = a^\ast, X_i = x_i, X_j=x_j, X_n=x_n | \textrm{subgraph formed by nodes 1,i,j,n})p_{ij}^t.$
\end{enumerate}
\end{frame}
